---
title: "5702_final_dmh"
output: html_document
---
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(lubridate)
library(ggpubr)
library(mi)
```
### data preprocessing
The original YoY_Seated_Diner_Data-2.csv file contains information about 8 country of 60 state and 60 cities. Here we only extract US state information 

There are many NA's but we decide to subset the data we need and then deal with NA's

```{r}
us_case = read.csv("./data/us-states.csv")
restaurant = read.csv("./data/YoY_Seated_Diner_Data-2.csv")
restaurant1 = restaurant %>% pivot_longer(cols = starts_with("X"),
                           names_to = "Date",
                           values_to = "Percentage")%>%
               mutate(Date = paste0("2020.",str_remove(Date,"X")))%>%
               mutate(Date = as.Date(chartr(old = ".", new = "-", Date)))
reopen = read.csv("./data/YoY_Reopened_Seated_Diner_Data.csv")
reopen1 = reopen %>% pivot_longer(cols = starts_with("X"),
                           names_to = "Date",
                           values_to = "Percentage")%>%
               mutate(Date = paste0("2020.",str_remove(Date,"X")))%>%
               mutate(Date = as.Date(chartr(old = ".", new = "-", Date)))
```
Firstly we want to see the US national trend of restaurant reservation compared to the same time last year.

```{r}
us_states = c('Alabama','Arizona','California','Colorado', 'Connecticut',
       'District of Columbia','Florida', 'Georgia', 'Hawaii', 'Illinois',
       'Indiana', 'Kansas', 'Kentucky', 'Louisiana',
       'Maryland', 'Massachusetts', 'Michigan',
       'Minnesota', 'Missouri', 'Nebraska', 'Nevada', 'New Jersey',
       'New Mexico', 'New York', 'North Carolina', 'Ohio', 'Oklahoma', 'Oregon',
       'Pennsylvania','Rhode Island', 'South Carolina', 'Tennessee', 
       'Texas', 'Utah', 'Virginia', 'Washington', 'Wisconsin')

restaurant_state = restaurant1[restaurant1$Type == 'state',]
restaurant_state = restaurant_state[restaurant_state$Name %in% us_states,]
US.nation.res = restaurant1[which(restaurant1$Type=="country" & restaurant1$Name=="United States"), ]

reopen_state = reopen1[reopen1$Type == 'state',]
reopen_state = reopen_state[reopen_state$Name %in% us_states,]
```

### Missing Values
```{r}
sum(is.na(restaurant_state))
```

By observing reopen dataset, we notice that each state contain consecutive missing values in the started from the beginning of April. And the first non-NA value started when this stated reopen for in-door dinning. Hence, states have more NA's reopen later, and may be more conservative for allowing reopenning.
From the number of NA, we can already know Texas, Florida reopen early for in-door dinning, New Jersey and New York reopen later for in-door dining
```{r}
reopen_na <- reopen_state %>%
  group_by(Name) %>%
  summarize(count_na = sum(is.na(Percentage))) %>%
  arrange(count_na)
reopen_na$Name = factor(reopen_na$Name,levels = reopen_na$Name)
ggplot(reopen_na)+
  geom_col(aes(x = Name,y = count_na))+
  ggtitle("Missing Value count plot")
```


```{r}
ggplot(US.nation.res) +
  geom_line(aes(x=Date, y=Percentage),color="#69b3a2") + 
  geom_hline(yintercept=0, color="orange", size=.5)+ 
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  theme_light()+
  ggtitle("Year-over-year Comparison of seated dinner in the US")
```
```{r}
US_case = read.csv("./data/us-states.csv")
US_case_by_date = US_case %>%
  mutate(date = as.Date(date))%>%
  group_by(date) %>%
  summarize(cumu_sum = sum(cases)) %>%
  mutate(daily_new = cumu_sum-lag(cumu_sum)) %>%
  drop_na()
US_case_by_date
```
```{r}
p1 <- ggplot() +
  geom_line(data = US.nation.res,aes(x=Date, y=Percentage),color="#69b3a2") + 
  geom_hline(yintercept=0, color="orange", size=.5)+ 
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  theme_light()+
  ggtitle("Year-over-year Comparison of seated dinner in the US")
```

```{r}
library("viridis")      
ggplot(restaurant_state, aes(x = Date, y = Name, fill = Percentage)) + 
  geom_tile()+
  scale_fill_viridis()
  
```

The heatmap above shows the number of seated diners by state from February until December 2020. We can see a clear drop in seated dinners across all states since mid-March.
Hawaii's restaurant industry was hit hardest and just had a slight recovery after November. District of Columbia also has a visibly severe impact on sitted dinner than other states. People in New York, New Jersey and Washinton seem to go for sitted dinner later. On the contrary,Alabama, Arizona, Florida, Oklahoma and Texas start eating out earlier. 

```{r}
p2<- ggplot() +
  geom_line(data = US_case_by_date,aes(x=date, y=daily_new),color="orange") + 
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  theme_light()+
  ggtitle("Daily COVID case in US")
```
```{r}
ggarrange(p1,p2,nrow=2)
```
From the year-over-year seated dinner rate plot, we can see the seated dinner hugely drop to -100% compared to the same time of last year, means all the seated dinner stopped nationwide by late March. The seated dinner rate started to increase starting from late April to June. After late June, the seated dinner rate became stable and showing an obious weekly pattern. By comparing the seated dinner rate and daily covid trend together, we can see that after states are reopen for seated dinner, the in-door dining rate did not change so much coresponding to change of COVID cases. Especially, after mid October, the daily covid case change hugely but people's willingness of in-door dining does not change so much, which might be dangerous.

After knowing the national situation, We want to know New York COVID situation was after reopen for in-door dinning.

```{r}
ggplot()
```



<!--chapter:end:5702_final_dmh.Rmd-->

---
title: "Airlines_FinalProject"
author: "Neha Pusarla"
date: "12/2/2020"
output: pdf_document
---

## R Markdown

Libraries 
```{r}
library("readxl")
library(ggplot2)
library(tidyverse)
```

Data Transformation 
```{r}
<<<<
data <- read_excel("../data/scheduled-canceled-operated-flights-jan2019-jul2020.xlsx")

year <- c(2019, 2019, 2019, 2019, 2019, 2019, 2019, 2019, 2019, 2019,
          2019, 2019, 2020, 2020, 2020, 2020, 2020, 2020, 2020)
my_data <- data.frame(year)

for (val in 2:length(data)){
  array <- data[val][[1]]
  array <- array[5:23]
  my_data <- cbind(my_data, array)
}

#rename the columns 
names <- c("year", "month", "Tot_Flights", "Alaska",
           "Allegiant", "American", "Delta", "Frontier", "Hawaiian", 
           "JetBlue", "Southwest", "Spirit", "United")
names(my_data) <- names 

#convert to numeric 
my_data$Tot_Flights <- as.numeric(my_data$Tot_Flights)
my_data$Alaska <- as.numeric(my_data$Alaska)
my_data$Allegiant <- as.numeric(my_data$Allegiant)
my_data$American <- as.numeric(my_data$American)
my_data$Delta <- as.numeric(my_data$Delta)
my_data$Frontier <- as.numeric(my_data$Frontier)
my_data$Hawaiian <- as.numeric(my_data$Hawaiian)
my_data$JetBlue <- as.numeric(my_data$JetBlue)
my_data$Southwest <- as.numeric(my_data$Southwest)
my_data$Spirit <- as.numeric(my_data$Spirit)
my_data$United <- as.numeric(my_data$United)

head(my_data)
```

Exploratory Analysis 
```{r}
#Main Question: How did Covid impact travel and transportation?

#difference between 2019 vs. 2020 in total flights 
tot_flights_2019 <- sum(my_data['Tot_Flights'][[1]][1:12])
tot_flights_2020 <- sum(my_data['Tot_Flights'][[1]][13:19])
tot_flight_data <- data.frame("year" = c("2019", "2020"), 
                              "flights" = c(tot_flights_2019, tot_flights_2020))
ggplot(tot_flight_data, aes(x = year, y = flights)) +
  geom_bar(stat = "identity", fill='dodgerblue1') +
  ggtitle("Bar Graph of Total Flights in 2019 vs. 2020") 
```

As seen from an overall perspective, there are a lot fewer flights taken in 2020 than in 2019, most likely due to the pandemic. Although, the 2019 data includes all months, while the 2020 data includes only the first 7 months, therefore, we will further our analysis using only the first 7 months of the 2019 data as well.

```{r}
tot_flights_2019 <- sum(my_data['Tot_Flights'][[1]][1:7])
tot_flights_2020 <- sum(my_data['Tot_Flights'][[1]][13:19])
tot_flight_data <- data.frame("year" = c("2019", "2020"), 
                              "flights" = c(tot_flights_2019, tot_flights_2020))
ggplot(tot_flight_data, aes(x = year, y = flights)) +
  geom_bar(stat = "identity", fill='dodgerblue1') +
  ggtitle("Bar Graph of Total Flights in First 7 Months 2019 vs. 2020") 
```


Using only the first 7 months of the 2019 data to compare its results more accuractely with the first 7 months of the 2020 data, there is still a clear distinctment between the number of flights taken in 2019 vs. 2020. But, since we are using all of our 2020 data, we cannot make any analyses on the effect of the pandemic on travel as travel restrictions began in March. Therefore, we are going to plot the pattern of flights taken throughout last year and this year to analyze the impact.

```{r}
data_2020 <- my_data[my_data$year==2020, ]

ggplot(my_data, aes(fct_inorder(month), Tot_Flights, col = year, group = year)) +
  geom_line() + 
  geom_point() + 
  ggtitle("Line Graph of Total Flights by Months") + 
  xlab("Month") + 
  ylab("Total Flights")
```

Using this graph, we can clearly see a big drop in flights from March to April of 2020. One can assume from this drop that the pandemic most likely played a role in the decrease of flights. Around May, the number of flights slowly start to increase, but not to the level that it was at the beginning of the year, as can be seen by the comparison to the flights taken in May of 2019. Even though flight frequency increased, it is still greatly decreased from the regular frequency of flights. Let us see how the pattern of the number of coronavirus cases in the United States affected the travel pattern. 

```{r}
df_jan_cases <-  df_case %>% 
  filter(date == "2020-01-29")
jan_cases <- sum(df_jan_cases$cases)
df_feb_cases <-  df_case %>% 
  filter(date == "2020-02-29")
feb_cases <- sum(df_feb_cases$cases) - sum(df_jan_cases$cases)
df_mar_cases <-  df_case %>% 
  filter(date == "2020-03-29")
mar_cases <- sum(df_mar_cases$cases) - sum(df_feb_cases$cases)
df_apr_cases <-  df_case %>% 
  filter(date == "2020-04-29")
apr_cases <- sum(df_apr_cases$cases) - sum(df_mar_cases$cases)
df_may_cases <-  df_case %>% 
  filter(date == "2020-05-29")
may_cases <- sum(df_may_cases$cases) - sum(df_apr_cases$cases)
df_jun_cases <-  df_case %>% 
  filter(date == "2020-06-29")
jun_cases <- sum(df_jun_cases$cases) - sum(df_may_cases$cases)
df_jul_cases <-  df_case %>% 
  filter(date == "2020-07-29") 
jul_cases <- sum(df_jul_cases$cases) - sum(df_jun_cases$cases)

month_cases <- c(jan_cases, feb_cases, mar_cases, apr_cases, may_cases, jun_cases, jul_cases)

data_2020 <- cbind(data_2020, month_cases)
tidydf <- data_2020 %>%
  mutate(cases =round(100*month_cases/month_cases[4], 2)) %>%
  mutate(flights = round(100*Tot_Flights/Tot_Flights[4], 2))

tidydf <- tidydf %>% select(month, cases, flights) %>%
  gather(key = "variable", value = "value", -month)


p1 <- ggplot() +
  geom_line(data = data_2020, aes(x=fct_inorder(month), y=Tot_Flights, group=1),color="#69b3a2") + 
  geom_point(size = 4) + 
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  theme_light()+
  ggtitle("Total Flights in 2020 in the US") + 
  xlab("Month") + 
  ylab("Total Flights")

p2<- ggplot() +
  geom_line(data = data_2020 , aes(x=fct_inorder(month), y=month_cases, group=1),color="orange") + 
  geom_point(size = 4) + 
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  theme_light()+
  ggtitle("COVID cases per Month in US") + 
  xlab("Month") + 
  ylab("Total Cases")

ggarrange(p1,p2,nrow=2)

```

This graph shows when the coronavirus hit the United States hard, and affected many different factors, one being the airline industry and travel. Although there were cases in January, the number of cases quickly ramped up from March to April, which correlates with the decline in travel. From the graph, it can be seen that around May, although cases were still increasing, travel started to increase again, as people started to most likely relax on restrictions. 

```{r}
#See how much percent each airline was affected 
percent_changes <- c() 
for (val in 4:(length(data_2020)-1)){
  array <- data_2020[val][[1]]
  percent_change <- 
    ((sum(array[1:3]) - sum(array[4:7])) / 
       sum(array[1:3])) * 100 
  percent_changes <- append(percent_changes, percent_change)
}

airlines <- c("Alaska","Allegiant", "American", "Delta", "Frontier", "Hawaiian", "JetBlue", "Southwest", "Spirit", "United")

ggplot(data.frame(airlines, percent_changes), 
       aes(reorder(airlines, -percent_changes), percent_changes)) +
  geom_bar(stat = "identity", fill='dodgerblue1') +
  ggtitle("Bar Graph of Percent Decrease for each Airline") + 
  xlab("Airline") + 
  ylab("Percent Decrease")
```

This data shows by how much percent the number of flights in
each airline decreased from the first couple months, pre-covid,
to the rest of the year, after the oncoming of the pandemic in
the United States. As shown by the graph, every single airline
was affected by the pandemic as they all have a positive percent
change, meaning that there was a decline in the number of flights
for each airline. The airlines that were affected the most were
JetBlue, Hawaiian, and United. These airlines had at least a 60%
decline in flights when the pandemic started.

```{r}
top_drops <- data_2020 %>%
  select(month, JetBlue, Hawaiian, United, Delta) %>%
  gather(key = "variable", value = "value", -month)
ggplot(top_drops, aes(fct_inorder(month), value, group = variable)) + 
  geom_line(aes(color = variable), size = 1) +
  geom_point() + 
  ggtitle("Top 4 Airline Brands Affected") + 
  xlab("Month") + 
  ylab("Flights")
```

This graph shows the time series data of the top four airline brands affected, taken from the previous graph. The trends stay consistent with the trends that we've been seeing in the previous graphs. 

Final Findings on Airline Data: 
The most revealing findings from this exploratory data analysis showed a couple of things. Firstly, there was a major decrease in flights from the year 2019 to the year 2020. Secondly, the number of flights mainly dropped in April of 2020, correlating with the time that the lockdown started and travel restrictions were placed everywhere. Looking at each airline, there was a decrease of the total number of flights taken for each airline, showing that almost every airline industry was affected by the pandemic. Some airlines had a decrease in flights by more than 60%, showing the intensity of the effect on the airline industry. More than half of their revenue was gone. Overall, in terms of the effect on travel during the coronavirus pandemic, there are clear patterns that show that travel was greatly affected. Airplanes were not a first choice method of transporation anymore, as people were not even travelling in the first place. This major decline in airline transportation and travel in and out of the United States, shows how greatly the pandemic played a role in the decline of the airline industry. 

<!--chapter:end:Airlines_Final_Project.Rmd-->

---
title: "covid_case"
author: "Run"
date: "12/1/2020"
output: html_document
---


load package 
```{r}
library(tidyverse)
library(choroplethr)
library(ggplot2)
library(USAboundaries)
library(plotly)
library(readxl)
```

us population estimation in 2019 from https://www.census.gov/data/tables/time-series/demo/popest/2010s-total-cities-and-towns.html 



###ppldata
```{r}
df_ppl <- read_excel("../data/nst-est2019-01.xlsx", range = c("A10:M60"),
                      col_names = c("geographic_area", "ppl2000","ppl2000est","0","1","2","3","4","5","6","7","8","ppl2019")) %>% 
  mutate(state = substring(geographic_area,2)) %>% 
  select("state","ppl2019")

```   

### us covid cases in states
```{r}
# prepare covid cases df
df_case = read.csv("../data/us-states.csv") 
df_total_cases <-  df_case %>% 
  filter(date == "2020-11-29") 

df_total_cases <-  df_total_cases %>% 
  merge(x = df_total_cases , y=USAboundaries :: state_codes, by.x="state", by.y = "state_name")
df_total_cases <-  df_total_cases %>% 
  merge(x = df_total_cases , y=df_ppl, by.x="state", by.y = "state") %>% 
  mutate(case_ppl = cases/ppl2019) %>% 
  mutate(case_ppl_log = log(cases/ppl2019)) %>%
  mutate(case_log = log(cases))

df_total_cases$case_ppl <-  round(df_total_cases$case_ppl, 3)

df_total_cases$hover <- with(df_total_cases, paste(state, '<br>', "Covid cases",cases,"<br>",
                                                   "Covid cases per capita",case_ppl))
                                                   

```


```{r}
# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
# specify some map projection/options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)


fig <- plot_geo(df_total_cases, locationmode = 'USA-states')
fig1 <- fig %>% add_trace(
    z = ~case_ppl, text = ~hover, locations = ~state_abbr,
    color = ~case_ppl*100, colors = 'Reds'
  )
fig1 <- fig1 %>% colorbar(title = "number cases per capita")
fig1 <- fig1 %>% layout(
    title = 'Positive cases of coronavirus per capita in the United States<br>(Hover for breakdown)',
    geo = g
  )

fig1


```

```{r}
fig2 <- fig %>% add_trace(
    z = ~cases, text = ~hover, locations = ~state_abbr,
    color = ~cases, colors = 'Reds'
  )
fig2 <- fig2 %>% colorbar(title = "number of cases")
fig2 <- fig2 %>% layout(
    title = 'Positive cases of coronavirus in the United States<br>(Hover for breakdown)',
    geo = g
  )

fig2

```

```{r}
df_total_cases %>%
  mutate(state = fct_reorder(state, desc(-cases))) %>%
  ggplot(aes(x = state, y = cases)) +
    geom_point(stat = "identity")+
    coord_flip() +
    ggtitle("Total Confirmed Covid19 Cases for Each State in The U.S.")
    
```



```{r}
#data process, us_cases1 is the overall us cases per day
us_cases1 <- df_case  %>% 
  count( date,wt = cases,name = "total_cases") %>% 
  mutate(date = as.Date(us_cases1$date)) %>% 
  mutate(new_cases = (c(1,diff(total_cases))))
```


### Covid19 Cases in Time Series
```{r}
ggplot(us_cases1,aes(x=date,y=total_cases)) +
  geom_line(color="#69b3a2") + 
  scale_x_date(date_breaks = "1 month", date_labels = "%b")+
  theme_light()+
  ggtitle("Cumulative Covid19 Cases in the US")
```



```{r}
ggplot(us_cases1,aes(date,new_cases)) +
  geom_line(color="#69b3a2") + 
  geom_smooth(method = "loess", se = FALSE,span=0.2, lwd = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")+
  theme_light()+
  ylab("daily increase")
  ggtitle("Daily New Covid19 Cases in the US")
```


```{r}


library(parcoords)
df_ppc_case <- df_total_cases 
rownames(df_ppc_case) <- df_ppc_case$state 
df_ppc_case <- df_ppc_case%>% 
  select('cases','case_ppl') %>% 
  rename(cases_per_capita = case_ppl)

parcoords(df_ppc_case,
          reorderable = T,
          brushMode = '1D-axes')

```

```{r}
df <- df_total_cases %>% 
  mutate(case_ppl_order = rank(-case_ppl)) %>% 
  mutate(sever_level = cut(case_ppl_order, 
                   breaks=c(0, 10, 20, 51), 
                   labels=c(3,2,1))) %>% 
  select("state","cases","case_ppl","sever_level")
```



```{r}
library(ggplot2)
library(GGally)
#set the value of alpha to 0.5
ggparcoord(df, columns=c(2:3), groupColumn = 4, alphaLines = .9, title = "Cases v.s. Cases Per Capita",scale = "uniminmax") +
   scale_color_manual(values=c("maroon","gray14","gray")) +
  theme_grey(16)
```





<!--chapter:end:covid_case.Rmd-->

---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(dplyr)
library(mi)
library(tidyverse)
library(forcats)
```


```{r}
hospital <- read.csv("../data/reported_hospital_utilization_20201130_2136.csv")
head(hospital)
```
This data is about hospital utilization by state

### Missing Values
```{r}
sum(sort(colSums(is.na(hospital)), decreasing = TRUE))
```
 
 As we see above, there are no missing values.
 
```{r}
staff_shortage <- c(1:7)
inpatient_bed <- c(1, 10:15)
prev_day <- c(1, 16:23)
staffed <- c(1, 24:29)
total <- c(1, 30:39)
inpatient_bed <- c(1, 40:51)
icu_bed <- c(1, 52:59)
staff_shortage_df <- hospital[,staff_shortage]
inpatient_bed_df <- hospital[,inpatient_bed]
prev_day_df <- hospital[,prev_day]
total_df <- hospital[,total]
inpatient_bed_df <- hospital[,inpatient_bed]
icu_bed_df <- hospital[,icu_bed]
```

```{r}
hospital[,60][1]
```

All data in this folder is recorded for the date above.




### Beds use for covid
```{r}
#testing
total_bed <- hospital$inpatient_beds

covid_bed <- hospital$inpatient_beds_used_covid
not_covid_bed <- hospital$inpatient_beds_used - covid_bed
empty_bed <- total_bed - hospital$inpatient_beds_used
state <- as.factor(hospital$state)

bed_usage <- data.frame(state, covid_bed, not_covid_bed, empty_bed)
bed_usage_tidy <- bed_usage %>% 
  mutate(covid_bed_perc = covid_bed/(covid_bed+not_covid_bed+empty_bed)) %>% 
  pivot_longer(cols = !c(state,covid_bed_perc), names_to = "bedtype", values_to = "count") %>% 
  mutate(state = fct_reorder(state, desc(-covid_bed_perc))) %>% 
  mutate(bedtype = factor(bedtype, level = c("covid_bed","empty_bed","not_covid_bed")))

  ggplot(bed_usage_tidy,aes(x = state , y = count, fill = (bedtype))) + 
    geom_bar(,stat = "identity", position = position_fill(reverse = TRUE)) + 
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = 'state') +
    coord_flip() +
    ylab("proportion") + 
    xlab("State") +
    theme(legend.position = "bottom") +
    ggtitle("Inpatient Bed Usage by State")
```


Through this visualization, we see that the majority of beds occupied for all states are not covid related. As of November 26, there are more empty beds than beds occupied by covid patients. However, that difference is not very large in many states that may have less beds overall. To build on this intuition, we will explore the hospital's own expressions about having a critical staffing shortage.

```{r}
#percent of hospitals that express critical staffing shortage for this day
staff_short_prop <- hospital$critical_staffing_shortage_today_yes / (hospital$critical_staffing_shortage_today_yes+hospital$critical_staffing_shortage_today_no)
staff_short_df <- data.frame(hospital$state, staff_short_prop)

ggplot(staff_short_df, aes(x=staff_short_prop)) + 
  geom_histogram(bins=12) +
  ggtitle("% Hospitals in State with Critical Staffing Shortage") +
  xlab("% of Hospitals in State")
```
From this histogram, we see that most states have over 10% of their hospitals reporting critical staffing shortage as of November 26. 
```{r}
staff_short_df
```


# County Policy
```{r}
county <- read.csv("../data/state_policy_updates_20201202_0721.csv")
county$policy_type <- county$policy_type
head(county)
```

### Missing Values
```{r}
sort(colSums(is.na(county)), decreasing = TRUE)
x <- missing_data.frame(county)
image(x)
```

We observe many policies do not include total phases. Understandably, state wide policy levels will have no values in county and zip code. From this observation, we will look at out data from both state and county point of view.

```{r}
unique(county$policy_type)
county
```

```{r}

```


```{r}
library(tidyverse)
county_policy <- county[county$policy_level=='county' & county$start_stop=='start',]
county_policy$policy_type <- as.factor(county_policy$policy_type)
ggplot(county_policy, aes(fct_rev(fct_infreq(policy_type)))) + 
  geom_bar() +
  coord_flip() +
  ggtitle("County Frequency of Policy Types") +
  xlab("Policy Types")
```
From our data, there are 64 different policies. However this bar chart shows that around half of the policies are only enacted in a few counties. Policies such as "Shelter in Place", "Food and Drink", "Outdoor Recreation" is very common.


### State

```{r}
state <- county[county$policy_level == "state",]
ggplot(state[state$start_stop=='start',], aes(fct_rev(fct_infreq(policy_type)))) + 
  geom_bar() +
  coord_flip() +
  ggtitle("State Frequency of Policy Types") +
  xlab("Policy Types")
```
When we look at only States, there is a slight change in the distribution. Entertainment is the most commonly enacted policy. Food and Drink as well as Non-Essential Business is still very common. We see that some of these policies are enforced by most states.

```{r}
state
```


<!--chapter:end:hospital_county.Rmd-->

