---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(dplyr)
library(mi)
library(tidyverse)
```


```{r}
hospital <- read.csv("Downloads/reported_hospital_utilization_20201130_2136.csv")
head(hospital)
```
This data is about hospital utilization by state

### Missing Values
```{r}
#sort(colSums(is.na(hospital)), decreasing = TRUE)
```
 
 As we see above, there are no missing values.

### Data Exploration
```{r}
staff_shortage <- c(1:7)
inpatient_bed <- c(1, 10:15)
prev_day <- c(1, 16:23)
staffed <- c(1, 24:29)
total <- c(1, 30:39)
inpatient_bed <- c(1, 40:51)
icu_bed <- c(1, 52:59)
staff_shortage_df <- hospital[,staff_shortage]
inpatient_bed_df <- hospital[,inpatient_bed]
prev_day_df <- hospital[,prev_day]
total_df <- hospital[,total]
inpatient_bed_df <- hospital[,inpatient_bed]
icu_bed_df <- hospital[,icu_bed]
```

```{r}
hospital[,60][1]
```

Day of recording all this information

Beds used for covid
```{r}
total_bed <- hospital$inpatient_beds

covid_bed <- hospital$inpatient_beds_used_covid
not_covid_bed <- hospital$inpatient_beds_used - covid_bed
empty_bed <- total_bed - hospital$inpatient_beds_used
state <- as.factor(hospital$state)

bed_usage <- data.frame(state, covid_bed, not_covid_bed, empty_bed)
bed_usage_tidy <- bed_usage %>%
  pivot_longer(cols = !state, names_to = "bedtype", values_to = "count")

bed_usage_tidy %>%
  ggplot(aes(fill = bedtype, x=fct_rev(state), y=count)) +
  geom_bar(stat='identity', position='fill') +
  coord_flip() +
  ylab("proportion") + 
  xlab("State") +
  scale_fill_discrete(guide=guide_legend(reverse=T)) +
  theme(legend.position = "bottom") +
  ggtitle("Bed Usage by State")

```
```{r}
bed_usage_tidy

```


```{r}
staff_short_prop <- hospital$critical_staffing_shortage_today_yes / (hospital$critical_staffing_shortage_today_yes+hospital$critical_staffing_shortage_today_no+hospital$critical_staffing_shortage_today_not_reported)

```



# County Policy
```{r}
county <- read.csv("Downloads/state_policy_updates_20201202_0721.csv")
county$policy_type <- county$policy_type
head(county)
```
### Missing Values
```{r}
sort(colSums(is.na(county)), decreasing = TRUE)
```
```{r}

x <- missing_data.frame(county)
image(x)
```

We observe that state wide policy levels will have no values in county.
How should we consider those values into our chart?

```{r}
unique(county$policy_type)
```

```{r}
county$policy_type <- as.factor(county$policy_type)
```


```{r}
library(tidyverse)
ggplot(county, aes(fct_rev(fct_infreq(policy_type)))) + 
  geom_bar() +
  coord_flip() +
  ggtitle("Frequency of Different Policy Types by county") +
  xlab("Policy Types")
```

### State
```{r}
state <- county[county$policy_level == "state",]
state
```
```{r}
ggplot(state[state$start_stop=='start',], aes(fct_rev(fct_infreq(policy_type)))) + 
  geom_bar() +
  coord_flip() +
  ggtitle("Start Policy Types by State") +
  xlab("Policy Types")
```
```{r}
CA <- state[state$state_id == 'NY',]
CA
```
which state got affected the most
